Verification of Stencil Computations through Spatial Specifications

### 2016-07-07 paper.tex
checking editing

The sum of two regions $r \texttt{+} r'$ specifies that an array is
read using the neighbourhood indices described by either $r$ or
$r'$; it acts as a disjunction on specifications. For
example, the following gives the specification of a five-point
stencil which is the sum of two compound \texttt{reflexive} and
\texttt{centered} regions in each dimension:
%%
\begin{minted}{fortran}
!= stencil centered(depth=1, dim=1)*reflexive(dim=2) + centered(depth=1, dim=2)*reflexive(dim=1) :: a
b(i,j) = -4*a(i,j) + a(i-1,j) + a(i+1,j)
                   + a(i,j-1) + a(i,j+1)
\end{minted}
%%
Here the left hand side of \texttt{+} says that when the second dimension
(induction variable $j$) is fixed at the origin, the first dimension
(induction variable $i$) accesses the immediate vicinity of the origin
(to depth of one). The right hand side of \texttt{+} is similar but the dimensions are reversed.
This reflects the symmetry under rotation of the five-point stencil.

@@

\subsection{Indexing schemes and generalisation}

@@

For each assignment statement whose left-hand side is an array
subscript on neighbourhood indices, a finite map is computed which
maps array variables to a set of vectors representing array
subscripts.  This finite map contains all array subscript expressions
which flow to this statement. More formally, a function
$\textsf{analyse}$ is applied to each statement in a program with the
following clause:
%
\begin{align*}
\hspace{-0.5em}
\begin{array}{lr}
\textsf{analyse}(v(\overline{e_1}) = e_2)
 := \qquad\qquad & \textit{where} \; \neigh(\overline{e_1}) \wedge \arrayTy(v)  \\[0.3em]
\multicolumn{2}{l}{\qquad \bigcup\{v' \mapsto \{
\textsf{schematic}(\bar{e})\} \mid v'(\bar{e}) \leftarrow \mathsf{flowsTo}(e_2),
  \arrayTy(v')\}}
\end{array}
\end{align*}
%
That is, we focus on assignments to an array subscript where the LHS 
indexing expression $\overline{e_1}$ is a neighbourhood index.  For
all array subscripts that flow to the right-hand side of this
statement, a finite map is constructed, mapping each array variable
to a set of indexing schemes for its subscripts, computed
with \textsf{schematic} (Defn.~\ref{def:schematic},
p.~\pageref{def:schematic}).
 Note that \textsf{schematic} is undefined if
$\bar{e}$ contains relative indices which are not neighbourhood
indices (\ie{}, these are not stencil computations we can handle).

For our example, the \textsf{analyse} function matches on
line 5, with the following set for $\textsf{flowsTo}(\texttt{a(i, j) + x +
  y})$:
%
\begin{align*}
\{\texttt{a(i-1, j)}, \texttt{a(i+1, j)}, \texttt{a(i, j-1)},
  \texttt{a(i, j+1)}, \texttt{a(i, j)}\}
\end{align*}
Subsequently the result of \textsf{analyse} on line 5 yields the map:
\begin{align*}
\texttt{a} \mapsto \{\vtwoh{-1}{0}, \vtwoh{1}{0},
          \vtwoh{0}{-1}, \vtwoh{0}{1}, \vtwoh{0}{0}\}
\end{align*}
%
If the LHS $\overline{e_1}$ contains non-$0$ neighour offsets 
then all RHS schemes are \emph{relativized} by
$\textsf{schematic}(\overline{e_1})$. For example, 
if \texttt{a(i+1) = b(i)} then relativisation of the RHS by the LHS
produces an analysis which is equivalent to the analysis of \texttt{a(i) = b(i-1)}.

@@

Let $U$ range over the finite maps computed by \textsf{analyse}.  For
each $x \in \mathsf{dom}(U)$, the algorithm 
constructs next a set of $n$-dimensional rectangles covering all contiguous groups of schemes
in $U(x)$.

@@

\begin{definition}The $\oplus$ operation on $\textsf{A}(\specDNF)$ 
is defined in a similar way to $\odot$, with the intermediate $\hat{\oplus}$ on
$\specDNF$ given by:
%%
\begin{equation*}
S \,\hat{\oplus}\, \epsilon = S \quad
\epsilon \,\hat{\oplus}\, S = S \quad
S \, \hat{\oplus} \, T = S \, \term{+} \, T
\end{equation*}
%%
Thus, if neither specification is empty, their $\oplus$ is just
the syntatic sum $\term{+}$. Then $\oplus$ is the lifting of
$\hat{\oplus}$ to boudns with the shape as $\odot$ in the previous
definition.
\end{definition}

\begin{lemma}[$\hat{\odot}$, $\hat{\oplus}$ soundness] $\forall S,
  T \in \specDNF$ where $S \neq \epsilon \wedge T \neq \epsilon$:
\begin{align*}
\interp{S \; \hat{\odot} \; T}_n & = \interp{S}_n \otimes \interp{T}_n
  \\
\interp{S \; \hat{\oplus} \; T}_n & = \interp{S}_n \cup \interp{T}_n
\end{align*}
\label{lem:alg-soundness}
\end{lemma}
%

\subsection{Checking\label{subsec:checking}}

In checking mode, the specification is verified with respect to the indexing
scheme of the actual access pattern. The mathematical model of the specification
developed in Section~\ref{sec:semantics} is used this process.

Checking procedure synthesises syntactic specification into its model and
generates an injection from indexing schemes to the multiplicity of observed
access patterns, this is formally defined in Figure~\ref{fig:consistency}.
Then, the consistency of the injection from the indexing schemes with the model
is checked.

It the multiplicity of the access pattern and the model is matched and are
$\mathit{once}$, then the problem is reduced to checking consistency in
models in which duplicate access patterns are allowed.

Next, a consistency predicate considers whether the stencil is an approximation.
Semantically, a lower bound is achieved if every indexing scheme in the
model generalises some relativised access pattern that means all cells
described in the specification are part of the access pattern. Conversely,
an upper bound means all observed access patterns can be generalised with
the indexing schemes of the model. Hence, the actual cells are a subset of the
specified ones. In combination an exact specification, \ie{} one without any
approximation modifiers, is one that is both a lower and an upper bound.

\begin{figure}
\input{consistency.tex}
\caption{Consistency function of a model against stencil computation}
\label{fig:consistency}
\end{figure}

We will consider an extended example of a stencil computation that is
consistent with its annotation to show how checking works. The following is an
example with a looser but still consistent version of the \textit{fivepoint}
stencil specification inferred previously in the
Subsection~\ref{subsec:inference}:
%
\begin{minted}[breakindent=8.1em]{fortran}
!= region fivepoint = centered(depth=1,dim=1) + centered(depth=1,dim=2)
!= stencil readOnce, fivepoint :: b
a(i,j) = b(i, j-1) + b(i,j) + b(i,j+1) + b(i-1,j) + b(i+1,j)
\end{minted}
%
\begin{enumerate}
\item From now on assignment on line three is referred as $L3$ for brevity.
  Verification is done by computing the following:
  $$check(\texttt{stencil readOnce, fivepoint},L3,b)$$
%
\item Generate model from the specification as described in
  Section~\ref{sec:semantics}. We refer to the indexing
  schemes within the model as set $m$.
%
\[ m= \{  \vtwo{-1}{\infty},
          \vtwo{0}{\infty},
          \vtwo{1}{\infty},
          \vtwo{\infty}{-1},
          \vtwo{\infty}{0},
        \vtwo{\infty}{1} \} \]
\begin{dmath*}
\interp{\texttt{stencil readOnce, fivepoint}}_2 =
  \{\textsf{once}(\textsf{exact}(m)) \}
\end{dmath*}
%
\item To evaluate the two predicates above, we need indexing schemes for the
  array in question. This is obtained by $modelise$ function. For brevity the
  set of indexing schemes is referred as $m'$.

  For the example above the observed model is as follows:
%
\[
m' = \{
    \vtwo{0}{-1},
    \vtwo{0}{0},
    \vtwo{0}{1},
    \vtwo{-1}{0},
    \vtwo{1}{0}
  \}
\]
\[modelise(L3,b) = \textsf{once}(m')\]
%
\item Now $cons$ predicate is applied with this model along side with the
  assignment on line three and array $b$.

  The predicate is reduced to
  $cons(\textsf{once}(\textsf{exact}(m)),\textsf{mult}(m'))$.
%
\item $cons$ predicate from the previous step evaluates to
  $$cons(\textsf{mult}(\textsf{exact}(m)),\textsf{once}(m')).$$
%
\item We further evaluate the predicate to
  \begin{dmath*}
    cons(\textsf{mult}(\textsf{up}(m)),\textsf{mult}(m')) \wedge
    cons(\textsf{mult}(\textsf{low}(m),\textsf{mult}(m')).
  \end{dmath*}
%
\item Below we present two sets of generalisations to show both the lower and
  the upper bound holds.

  The generalisations for the upper bound:
  \begin{align*}
     \vtwoh{0}{-1} & \preceq_2 \vtwoh{0}{\infty} \\
     \vtwoh{0}{0} & \preceq_2 \vtwoh{0}{\infty} \\
     \vtwoh{0}{1} & \preceq_2 \vtwoh{0}{\infty} \\
     \vtwoh{-1}{0} & \preceq_2 \vtwoh{\infty}{0} \\
     \vtwoh{1}{0} & \preceq_2 \vtwoh{\infty}{0}
  \end{align*}

  The generalisations for the lower bound:
  \begin{align*}
     \vtwoh{-1}{0} & \preceq_2 \vtwoh{-1}{\infty} \\
     \vtwoh{0}{0} & \preceq_2 \vtwoh{0}{\infty} \\
     \vtwoh{1}{0} & \preceq_2 \vtwoh{1}{\infty}  \\
     \vtwoh{0}{-1} & \preceq_2 \vtwoh{\infty}{-1} \\
     \vtwoh{0}{0} & \preceq_2 \vtwoh{\infty}{0} \\
     \vtwoh{0}{1} & \preceq_2 \vtwoh{\infty}{1}
  \end{align*}

  Since both predicates are satisfied the array accesses specified on line
  three, the stencil computation is consistent with the \texttt{fivepoint}
  specification.
\end{enumerate}

Now when $L3$ is modified with an additional term \texttt{b(i+2,j+2)},
the new assignment $L3'$ becomes:
%
\begin{minted}[firstnumber=3]{fortran}
a(i,j) = b(i, j-1) + b(i,j) + b(i,j+1) + b(i-1,j) + b(i+1,j) + b(i+2,j+2)
\end{minted}
%
In the first six steps, evaluation proceeds the same way except for step three.
In the third step, result of $\mathit{modelise}(L3',b)$ will have the extra term
$\vtwoh{2}{2}$. In the seventh step, since the model has not changed and the
indexing schemes are only augmented with a new element, the generalisations for
lower bound still hold. However, there are no elements in the model, $m$, that
can generalise $\vtwoh{2}{2}$, hence the model fails to hold as an upper bound.

Note that if the specification were as follows:
%
\begin{minted}{fortran}
!= stencil atLeast, readOnce, fivepoint :: b
\end{minted}
%
The consistency predicate would have held because for the upper bound
$\mathit{cons}(\textsf{mult}(\textsf{up}(m)),\textsf{mult}(m'))$ would not have
been needed. This semantically makes sense since the region $\texttt{fivepoint}$
is indeed contained in the regions touched by $b$.

\subsubsection{A note on stencil approximations}

>>>>

The sum of two regions $r \texttt{+} r'$ specifies that an array is
read using the neighbourhood indices described by either $r$ or
$r'$; it acts as a disjunction on specifications. For
example, the following gives the specification of a five-point
stencil which is the sum of two compound \texttt{reflexive} and
\texttt{centered} regions in each dimension:
%%
\begin{minted}{fortran}
!= stencil centered(depth=1, dim=1)*reflexive(dim=2) + centered(depth=1, dim=2)*reflexive(dim=1) :: a
b(i,j) = -4*a(i,j) + a(i-1,j) + a(i+1,j)
                   + a(i,j-1) + a(i,j+1)
\end{minted}
%%
Here the left-hand side of \texttt{+} says that when the second dimension
(induction variable $j$) is fixed at the origin, the first dimension
(induction variable $i$) accesses the immediate vicinity of the origin
(to depth of one). The right hand side of \texttt{+} is similar but the dimensions are reversed.
This reflects the symmetry under rotation of the five-point stencil.

@@

\subsection{Indexing schemes and generalisation}
\label{subsec:schemes-and-gen}

@@

For each assignment statement whose left-hand side is an array
subscript on neighbourhood indices, a finite map is computed which
maps array variables to a set of vectors representing array
subscripts.  This finite map contains all array subscript expressions
which flow to this statement. More formally, a function
$\textit{analyse}$ is applied to each statement in a program with the
following clause:
%
\begin{align*}
\hspace{-0.5em}
\begin{array}{lr}
\textit{analyse}(v(\overline{e_1}) = e_2)
 := \qquad\qquad & \textit{where} \; \neigh(\overline{e_1}) \wedge \arrayTy(v)  \\[0.3em]
\multicolumn{2}{l}{\qquad \bigcup\{v' \mapsto \{
\textsf{schematic}(\bar{e})\} \mid v'(\bar{e}) \leftarrow \mathsf{flowsTo}(e_2),
  \arrayTy(v')\}}
\end{array}
\end{align*}
%
That is, we focus on assignments to an array subscript where the LHS 
indexing expression $\overline{e_1}$ is a neighbourhood index.  For
all array subscripts that flow to the right-hand side of this
statement, a finite map is constructed, mapping each array variable
to a set of indexing schemes for its subscripts, computed
with \textsf{schematic} (Defn.~\ref{def:schematic},
p.~\pageref{def:schematic}).
 Note that \textsf{schematic} is undefined if
$\bar{e}$ contains relative indices which are not neighbourhood
indices (\ie{}, these are not stencil computations we can handle).

For our example, the \textit{analyse} function matches on
line 5, with the following set for $\textsf{flowsTo}(\texttt{a(i, j) + x +
  y})$:
%
\begin{align*}
\{\texttt{a(i-1, j)}, \texttt{a(i+1, j)}, \texttt{a(i, j-1)},
  \texttt{a(i, j+1)}, \texttt{a(i, j)}\}
\end{align*}
Subsequently the result of \textit{analyse} on line 5 yields the map:
\begin{align*}
\texttt{a} \mapsto \{\vtwoh{-1}{0}, \vtwoh{1}{0},
          \vtwoh{0}{-1}, \vtwoh{0}{1}, \vtwoh{0}{0}\}
\end{align*}
%
If the LHS $\overline{e_1}$ contains non-$0$ neighour offsets 
then all RHS schemes are \emph{relativized} by
$\textsf{schematic}(\overline{e_1})$. For example, 
if \texttt{a(i+1) = b(i)} then relativisation of the RHS by the LHS
produces an analysis which is equivalent to the analysis of \texttt{a(i) = b(i-1)}.

@@

Let $U$ range over the finite maps computed by \textit{analyse}.  For
each $x \in \mathsf{dom}(U)$, the algorithm 
constructs next a set of $n$-dimensional rectangles covering all contiguous groups of schemes
in $U(x)$.

@@

\begin{definition}The $\oplus$ operation on $\textsf{A}(\specDNF)$ 
is defined in a similar way to $\odot$, with the intermediate $\hat{\oplus}$ on
$\specDNF$ given by:
%%
\begin{equation*}
S \,\hat{\oplus}\, \epsilon = S \quad
\epsilon \,\hat{\oplus}\, S = S \quad
S \, \hat{\oplus} \, T = S \, \term{+} \, T
\end{equation*}
%%
Thus, if neither specification is empty, their $\hat{\oplus}$ is just
the syntatic sum $\term{+}$. Then $\oplus$ is the lifting of
$\hat{\oplus}$ to \textsf{Approx} with the shape as $\odot$ in the previous
definition.
\end{definition}

\begin{lemma}[$\hat{\odot}$, $\hat{\oplus}$ soundness] $\forall S,
  T \in \specDNF$ where $S \neq \epsilon \wedge T \neq \epsilon$:
\begin{align*}
\interp{S \; \hat{\odot} \; T}_n = \interp{S}_n \otimes \interp{T}_n
  \qquad
\interp{S \; \hat{\oplus} \; T}_n = \interp{S}_n \cup \interp{T}_n
\end{align*}
\label{lem:alg-soundness}
\vspace{-2em}
\end{lemma}
%
\noindent
Thus our model validates the specification synthesis process which concludes the
inference procedure. For our example, we infer and then synthesise the
5-point specification:
%%
\begin{minted}[linenos=false]{fortran}
!= stencil centered(depth=1, dim=1)*reflexive(dim=2) + centered(depth=1, dim=2)*reflexive(dim=1) :: a
\end{minted}
%%

\subsection{Specification checking\label{subsec:checking}}

In checking mode, the access pattern of a stencil computation in the
source language is verified against a specification. The model of
Section~\ref{sec:semantics} again aids this process.  Checking
proceeds by generating a model $M$ from a specification and generating a set $N$ of indexing
schemes from the source code. The source code is then valid with
respect to the specification if $N$ is \emph{consistent} with $M$. The
notion of consistency is built out of index scheme 
\emph{generalisation} (Sect.~\ref{subsec:schemes-and-gen}). 
Consistency is defined in Figure~\ref{fig:consistency},
but we first set up some intermediate definitions before unpacking its definition.

\begin{definition}[Checking]
Let \texttt{a} be an array variable and let
$S \in \textit{spec}$ be a specification with
the \textit{specDec} \texttt{!= stencil
  $S$ :: a} in our syntax.
For a array subscript
assignment statement $\textit{assg} = v(\overline{e_1}) = e_2$ in our source language
(with all the appropriate data flow information in place)
where $e_2$ has at most $n$-dimensional array access, 
then the assignment conforms to the specification $S$ if 
the predicate $\mathit{check}_n(S, \textit{assg}, \texttt{a})$ holds, 
defined:
\begin{align*}
& \mathit{check_n}(S, \textit{assg}, x) =
  \mathit{cons_n}(\interp{S}_n, \, \textit{modelise}(\textit{assg},
  x)) \\
& \mathit{modelise} : \textit{statement} \times \textit{variable}
  \rightarrow \textsf{Mult}((\domainVal)^n) \\ 
& \mathit{modelise} (\textit{assg}, x) =
  \begin{cases}
    \textsf{once}(\mathit{analyse}(\textit{assg})(x)) & \textit{unique}(\textit{assg},x) \\
    \textsf{mult}(\mathit{analyse}(\textit{assg})(x)) & otherwise
  \end{cases}
\end{align*}
Thus, checking proceeds by reusing the \textit{analyse} function of
inference (Section~\ref{subsec:inference}) to extract indexing
schemes for stencil computations flowing to an assignment. 
This is then augmented with multiplicity information
by $\mathit{modelise}$ where $\mathit{unique}(\textit{assg},x)$ is a
 predicate indicating whether indices to array $x$ in an
assignment $\textit{assg}$ are unique or not. 
\end{definition}

Subsequently the binary predicate for consistency, at the heart of checking
is defined over the sets: 
%
\begin{equation*}
\textit{cons}_n \subseteq \textsf{Mult}(\textsf{Approx}(\domainVal{}^n))
\times \textsf{Mult}(\domainVal{}^N)
\end{equation*}
%
where first component is a model of a specification,
and the second component is a set of indexing schemes generated from
the array subscripts in a program, with added multiplicity
information. Giving \textsf{Approx} information to the
index schemes inferred from the program is not necessary, nor
desirable-- we have an exact set of schemes. We refer to the first
component of the predicate as \emph{spec model} and the second as \emph{code model}.

\begin{definition}
Consistency is defined in Figure~\ref{fig:consistency}. We let
$M_s, M_c$ range over elements of $\domainVal{}^n$ where $M_s$ comes
from the spec model and $M_c$ comes from \textit{modelise} on the
source code. Let $N_c$ range over elements of
$\textsf{Mult}(\domainVal{}^n)$ coming from the code model, and $A_s$
over elements of $\textsf{Mult}(\textsf{Approx}(\domainVal{}^n))$ from the spec model.

The first equation states that, if the multiplicity of the spec and code
model are both $\mathsf{once}$, then the problem is reduced to checking consistency of
models in which non-unique indexing is allowed, labelled with
$\mathsf{mult}$.

The second equation considers the case where the specification is
an \term{atMost} upper-bound approximation (labelled with \textsf{up}
as shorthand for \textsf{upper} here). 
An upper bound specification means that for every scheme in the code
model, modelling a source-level index, $\forall u \in M_c$ there exists
a generalising scheme in the spec model $\exists v \in M$, \ie{}, $u
\preceq_n v$. That is, every source-level index is generalised by the model.

The third equation is the converse of an \term{atLeast},
lower approximation (labelled \textsf{low} as shorthand for
\textsf{lower}). Consistency is dual to the above: 
for every scheme in the spec model $\forall v \in M_s$ there
exists a scheme in the code model $\exists u \in M_c$ such that $u
\preceq_n v$. That is, every model scheme is generalised by source indexing.

These two are combined for consistency on an \textsf{exact}
specification (the fourth equation) such that the spec model
provides both a consistent lower and consistent upper bound on
the source code. Thus, for every scheme in the code model there
is a scheme in the spec model that generalises it, and for every
scheme in the spec model there is a scheme in the source model that
specialises it.

Lastly, if the spec model has both an upper and lower bound
consistency is defined inductively similar to the \textsf{exact} case.
\end{definition}

\begin{figure}
\input{consistency.tex}
\caption{Consistency between a model and indexing schemes from a stencil computation}
\label{fig:consistency}
\vspace{-0.9em}
\end{figure}

\begin{example}
We give an extended example of verifying a stencil computation against
its specification. We consider a more general, but still consistent, version of the \textit{five-point}
stencil specification inferred previously in Section~\ref{subsec:inference}:
%
\begin{minted}[breakindent=8.1em]{fortran}
!= region fivepoint = centered(depth=1,dim=1) + centered(depth=1,dim=2)
!= stencil readOnce, fivepoint :: b
a(i,j) = b(i,j-1) + b(i,j) + b(i,j+1) + b(i-1,j) + b(i+1,j)
\end{minted}
%
We refer to the assignment on line three as $l3$ for brevity.

Verification of the code against the spec is then given by 
$check(({\textcolor{darkgreen}{\texttt{readOnce, fivepoint}}}),l3,\texttt{b})$. 
We explain the steps.
\begin{enumerate}  
%
\item Generate the model from the specification as in
  Section~\ref{sec:semantics}. 
%
\begin{align*}
 M_s= \{  \vtwo{-1}{\infty},
          \vtwo{0}{\infty},
          \vtwo{1}{\infty},
          \vtwo{\infty}{-1},
          \vtwo{\infty}{0},
        \vtwo{\infty}{1} \}  \\[0.4em]
\interp{{\textcolor{darkgreen}{\texttt{readOnce, fivepoint}}}}_2 =
  \textsf{once}(\textsf{exact} (M_s))
\end{align*}
%
\item Obtained the code model via $\textit{modelise}$, giving:
%
\begin{align*}
M_c = \{
    \vtwo{0}{-1},
    \vtwo{0}{0},
    \vtwo{0}{1},
    \vtwo{-1}{0},
    \vtwo{1}{0}
  \}
\\[0.4em]
\textit{modelise}(l3, \texttt{b}) = \textsf{once} (M_c)
\end{align*}
%
\item Consistency $cons_2(\textsf{once}(\textsf{exact}(M_s)),
  \textsf{once}(M_c))$ holds when 
  $\textit{cons}_2(\textsf{mult}(\textsf{exact}(M_s)),\textsf{mult}(M_c))$ (Fig.~\ref{fig:consistency}) 
  which holds when: 
  \begin{align*}
     & \textit{cons}_2(\textsf{mult}(\textsf{up}(M_s)),\textsf{mult}(M_c)) \\
   \wedge \; & \textit{cons}_2(\textsf{mult}(\textsf{low}(M_s),\textsf{mult}(M_c))
  \end{align*}
%
\item This condition holds by the following sets of scheme
  generalisations: for the upper bound, $\forall u \in M_c,
  \exists v \in M_s . u \preceq_2 v$: 
  \begin{align*}
    \begin{array}{l}
     \vtwoh{0}{-1} \preceq_2 \vtwoh{0}{\infty} \quad
     \vtwoh{0}{0} \preceq_2 \vtwoh{0}{\infty} \quad
     \vtwoh{0}{1} \preceq_2 \vtwoh{0}{\infty} \\[0.5em]
     \vtwoh{-1}{0} \preceq_2 \vtwoh{\infty}{0} \quad
     \vtwoh{1}{0} \preceq_2 \vtwoh{\infty}{0}
     \end{array}
  \end{align*}
  % 
  and for the lower bound: $\forall v \in M_s, \exists u \in M_c . u
  \preceq_n v$: 
  \begin{align*}
    \begin{array}{l}
     \vtwoh{-1}{0} \preceq_2 \vtwoh{-1}{\infty} \quad
     \vtwoh{0}{0} \preceq_2 \vtwoh{0}{\infty} \quad
     \vtwoh{1}{0} \preceq_2 \vtwoh{1}{\infty}  \\
     \vtwoh{0}{-1}  \preceq_2 \vtwoh{\infty}{-1} \quad
     \vtwoh{0}{0}  \preceq_2 \vtwoh{\infty}{0} \quad
     \vtwoh{0}{1}  \preceq_2 \vtwoh{\infty}{1}
     \end{array}
  \end{align*}

  Since both the consistency predicate is satisfied, the array
  accesses flowing to line three in the stencil computation are valid
  with respect to the {\textcolor{darkgreen}{\texttt{fivepoint}}} specification.
\end{enumerate}
\noindent
Now when $L3$ is modified with an additional term \texttt{b(i+2,j+2)},
the new assignment $L3'$ becomes:
%
\begin{minted}[firstnumber=3]{fortran}
a(i,j) = b(i, j-1) + b(i,j) + b(i,j+1) + b(i-1,j) + b(i+1,j) + b(i+2,j+2)
\end{minted}
%
In the first six steps, evaluation proceeds the same way except for step three.
In the third step, result of $\mathit{modelise}(L3',b)$ will have the extra term
$\vtwoh{2}{2}$. In the seventh step, since the model has not changed and the
indexing schemes are only augmented with a new element, the generalisations for
lower bound still hold. However, there are no elements in the model, $m$, that
can generalise $\vtwoh{2}{2}$, hence the model fails to hold as an upper bound.

Note that if the specification were as follows:
%
\begin{minted}{fortran}
!= stencil atLeast, readOnce, fivepoint :: b
\end{minted}
%
The consistency predicate would have held because for the upper bound
$\mathit{cons}(\textsf{mult}(\textsf{up}(m)),\textsf{mult}(m'))$ would not have
been needed. This semantically makes sense since the region $\texttt{fivepoint}$
is indeed contained in the regions touched by $b$.
\end{example}
